## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
# if (logical statement) {
#  do something
# } else {
#  do something else
#}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10

if (a <= 5) {
  print(a)
}else {
  print("it's more than 5")
}
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 15

if (a<= 5){
  print('small')
} else if (a > 5 & a <= 10) {
  print('medium')
} else if (a > 10 | a == 7) {
  print('large')
} else {
  print("HUGE")
}
```

Neat! We can also nest conditionals just like we do in Python.

```{r}
a <- 5

if (a / 2 !=0) {
  if (a == 5){
    print('a is 5')
  }
}
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5) { # 1:5 is a list of range 1 - 5, inclusive.
  # Paste is a concatenation function & always adds space between elements
  print(paste(value, 'is greater than', value - 1))
  
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <- c(1,1,1,1,1,1)

#Change the first, third, and 6th elements to 6
for (i in c(1,3,6)) {
  a_vector[i] <- 6
}

a_vector
```

We can loop over more than just lists:

```{r}
mat <- matrix(1,nrow = 2, ncol = 2)

for (row in 1:2) {
  for (col in 1:2){
    mat[row, col] <- 0
    print(mat)
  }
}

for (row in 1:2) {
  for (col in 1:2){
    mat[col, row] <- 6
    print(mat)
  }
}
```

Like Python, we have while loops in R:

```{r}
b <- 0

while (b < 10) {
  # increase by two
  b <- b + 2
  print(b)
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
}
```

An example:

```{r}
# Start with a vector of temperatures
test_temp <- c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173557, 21.048836, 18.769735)

# declare function to convert from C to F:
c_to_f <- function(temp_c){
  temp_f <- temp_c * (9/5) + 32
  return(temp_f)
}

print(c_to_f(28))

```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
test_temp_f <- c_to_f(test_temp)  # maps over vector
test_temp_f
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,2,2), nrow = 2)
test_mat

# 1 is for rows, 2 is for columns
apply(test_mat, 1, sum)
apply(test_mat, 2, sum)

```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix(c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173557, 21.048836, 18.769735), nrow = 2)
test_temp_mat

apply(test_temp_mat, 1, c_to_f)
test_temp_mat # apply does NOT change original matrix!

```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
    #1: Remove the minimum

    # Task: 
    # Given an array of integers, remove the smallest value. Do not mutate the original array/list. If there are multiple elements with the same value, remove the one with a lower index. If you get an empty array/list, return an empty array/list.

    # Don't change the order of the elements that are left.

    remove_smallest <- function(numbers){
      if (length(numbers) == 0){return(numbers)
      }else {
        return(numbers[-match(min(numbers), numbers)[1]])
      }
    }
    remove_smallest(c(3,3,5,75,2,2,4,6,1,4,6,9))
    ```

    ```{r}
    #2: Breaking chocolate problem

    # Your task is to split the chocolate bar of given dimension n x m into small squares. Each square is of size 1x1 and unbreakable. Implement a function that will return minimum number of breaks needed.

    # For example if you are given a chocolate bar of size 2 x 1 you can split it to single squares in just one break, but for size 3 x 1 you must do two breaks.

    # If input data is invalid you should return 0 (as in no breaks are needed if we do not have any chocolate to split). Input will always be a non-negative integer.

    break_chocolate <- function (n, m) {
      # test for zero
      # n * m - 1
      
      if (n ==0 | m == 0){return(0)}
      return(n*m-1)
    }

    break_chocolate(4,9)
    ```

    ```{r}

    #3: Testing 1-2-3
    # Your team is writing a fancy new text editor and you've been tasked with implementing the line numbering.

    # Write a function which takes a list of strings and returns each line prepended by the correct number.

    # The numbering starts at 1. The format is n: string. Notice the colon and space in between.

    number <- function(lines) {
      
      i <- 1
      result <- c()
      for (line in lines) {
        result[i] <- paste(i,': ', line, sep = '')
        i <- i + 1
      }
      return(result)
    }

    number(c('a', 'b', 'c', 'I want a cookie'))
    ```

    ```{r}
    #4: Count the divisors of a number

    # Count the number of divisors of a positive integer n.
    divisors <- function(n) {
      divs <- c(n)
      i <- 2
      for (num in 1:n){
        if (n %% num == 0){
          divs[i] = num
          i <- i + 1
        }
      }
      # return(divs)
      return((length(divs) - 1))
    }

    divisors(50)
    ```

    ```{r}
    #5: Small enough? - Beginner

    # You will be given an array and a limit value. You must check that all values in the array are below or equal to the limit value. If they are, return true. Else, return false.

    # This was boring.
    small_enough <- function(vector, limit) {
      return(max(vector) <= limit)
    }

    small_enough(c(2,1,4,6,2,4,5,3,20), 12)
    small_enough(c(2,1,4,6,2,4,5,3), 12)
    ```

    ```         
    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}

    fizzbuzz <- function(num){
      if (num %% 3 == 0){
        if (num %% 5 == 0){
          return("fizzbuzz")
        }else {
          return("fizz")
        }
      }else if (num %% 5 == 0){
        return('buzz')
      }
      return("not divisible by 3 or 5")
    }

    fizzbuzz(10)
    fizzbuzz(3)
    fizzbuzz(6)
    fizzbuzz(15)


    for (number in 1:20){
      print(paste(number, 'is', fizzbuzz(number)))
    }

    # fizzbuzz(5)

    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.
    marioStairs <- function(steps){
      num_x <- 0
      while (steps > 0 ){
        num_x <- num_x + 1
        i <- 0
        level <- ''
        while (num_x > i){
          level <- paste(level,"X",sep = '')
          i <- i + 1
        }
        steps <- steps - 1
        print(level)
      } 
    }

    marioStairs(7)
    ```
